---
title: "Workshop4_Spatial_Mapping.qmd"
author: "Alan Tavares"
format: html
editor: visual
---

## Setup

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(sf)      # simple features for spatial data
# library(terra) # for raster data (not used in this basic workflow)
# library(tmap)  # optional - only if it installed successfully
```

## Load the copepod data

```{r}
#| label: load-data

dat <- read_csv("copepods_raw.csv")
```

## Data exploration

First look at the raw coordinates, coloured by species richness.

```{r}
#| label: explore-coords
#| fig-width: 7
#| fig-height: 5

ggplot(dat) +
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point() +
  scale_color_viridis_c(name = "Richness") +
  labs(
    title = "Copepod sample locations",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal()
```

Check richness against latitude for any unusual patterns.

```{r}
#| label: explore-richness
#| fig-width: 7
#| fig-height: 5

ggplot(dat, aes(x = latitude, y = richness_raw)) +
  stat_smooth() +
  geom_point(alpha = 0.5) +
  labs(
    title = "Copepod richness by latitude",
    x = "Latitude", y = "Species richness"
  ) +
  theme_minimal()
```

## Convert to a spatial (sf) object

```{r}
#| label: make-sf

sdat <- st_as_sf(dat, coords = c("longitude", "latitude"), crs = 4326)
sdat
```

## Check the coordinate reference system

```{r}
#| label: check-crs

st_crs(sdat)$Name
```

## Basic spatial plot

```{r}
#| label: basic-plot
#| fig-width: 7
#| fig-height: 5

plot(sdat["richness_raw"])
```

## Thematic map with ggplot2 (geom_sf)

This achieves the same result as `tmap`, using `ggplot2`'s built-in spatial support.

```{r}
#| label: ggplot-map
#| fig-width: 7
#| fig-height: 5.5

ggplot(sdat) +
  geom_sf(aes(color = richness_raw), size = 1.5, alpha = 0.8) +
  scale_color_viridis_c(name = "Species\nrichness") +
  labs(
    title = "Copepod species richness",
    subtitle = "AusCPR sampling locations",
    caption = "Data: AusCPR (modified for teaching purposes)"
  ) +
  theme_minimal()
```

## Adding polygon layers (Australia + continental shelf)

```{r}
#| label: load-shapefiles

aus <- st_read("Aussie/Aussie.shp")
shelf <- st_read("aus_shelf/aus_shelf.shp")
```

```{r}
#| label: layered-map
#| fig-width: 7
#| fig-height: 5.5

ggplot() +
  geom_sf(data = shelf, fill = "grey85", color = "grey60") +
  geom_sf(data = aus, fill = "grey40", color = "grey20") +
  geom_sf(data = sdat, aes(color = richness_raw), size = 1.2, alpha = 0.8) +
  scale_color_viridis_c(name = "Species\nrichness") +
  labs(
    title = "Copepod richness relative to the Australian coastline",
    subtitle = "Continental shelf shown in light grey",
    caption = "Data: AusCPR (modified for teaching purposes)"
  ) +
  theme_minimal()
```

## Optional: tmap version

Only run this section if `tmap` installed successfully.

```{r}
#| label: tmap-version
#| eval: false

library(tmap)

tm_shape(shelf, bbox = sdat) +
  tm_polygons() +
  tm_shape(aus) +
  tm_polygons() +
  tm_shape(sdat) +
  tm_dots(col = "richness_raw", palette = "viridis", title = "Richness")
```

## Summary

This workshop demonstrated:

- Loading and exploring copepod species richness data
- Converting tabular coordinates into a spatially referenced `sf` object
- Understanding coordinate reference systems (CRS)
- Building thematic maps combining point data with polygon layers, using `ggplot2::geom_sf()` as a reliable alternative to `tmap`
